---
title: "Decision tree analysis"
format: html
---

Load the data and split out days of week into separate boolean columns to be able to treat each as a separate variable.  Also, remove multi-campus workshops as their attendance numbers are not representative.

```{r message=FALSE}
library(tidyverse)
library(rpart)
library(rpart.plot)

df <- read_csv("our-workshops.csv", col_types="cDiflfiff") %>%
  filter(!multi_campus)

for (day in c("Mon", "Tue", "Wed", "Thu", "Fri")) {
  df[[day]] <- grepl(day, df$days_of_week)
}
```

For starters, just model the entire dataset.

```{r}
tree <- rpart(
  num_students ~ (
    modality +
    scheduling +
    week_in_quarter +
    Mon + Tue + Wed + Thu + Fri +
    time_of_day
  ),
  data=df
)

rpart.plot(tree, extra=1)
```
So higher attendance is associated with being online and being a single session in the morning.

But before we get too excited, how good is this model?  Pretty bad.  Let's calculate a few statistics:

```{r}
median = median(df$num_students)
rmse = sqrt(mean((predict(tree, df) - df$num_students)^2))
r2 = cor(predict(tree, df), df$num_students)^2
```

The RMSE is `r round(rmse, 0)`, which is incredibly large given that the median number of students is `r round(median, 0)`.  $R^2$ is only `r round(r2, 2)`.

Variable importance is as follows (these are percentages):

```{r}
importance <- round(
  tree$variable.importance/sum(tree$variable.importance)*100,
  0
)
importance
```

The top four variables (`r names(importance)[1]`, `r names(importance)[2]`, `r names(importance)[3]`, and `r names(importance)[4]`) are roughly equally important and account for `r importance[1]+importance[2]+importance[3]+importance[4]`% of the variability.
